WORKING TITLE: The adaptive value of green nest material across birds: A systematic review and meta-analysis

### 2025-08-16 manuscript_GNM.qmd
Added another sensitivity analysis without non-aromatic and aromatic comparison

Nest and nest-building are fundamental across a diverse range of animals, including arthropods (e.g., termites, spiders), mammals (e.g., platypus, rodents), and perhaps more famously, birds. Nests provide structural support for eggs, parents, and offspring, and protect them against predators and environmental pressures, and are thus key to reproductive success and survival [@hansell2000; @mainwaring2014; @deeming2023]. While most birds use organic plant materials such as twigs, branches, grass, and moss, some incorporate inorganic items, including mud, stones, feathers, snake skins and even anthropogenic materials like plastic and cigarette butts [@sergio2011; @suárez-rodríguez2013; @jagiello2023; @sheard2024]. These are considered non-structural additions and suggest that nests are more than just physical shelters; they can be complex tools shaped by both natural and sexual selection for multiple functions. Indeed, nests can play a key role in sexual display, and features such as size have been shown to act as sexually selected signals of parental quality in several bird species [@moreno1994; @hoi1996], even interspecifically [@soler1995]. Nest-building behaviour is increasingly recognized as an expression of the extended phenotype (reviewed by @moreno2012), and phylogeny has been shown to be an important factor explaining both mean and within-species variation in nest morphology [@perez2023]. Yet, despite hundreds of scientific articles studying nests and nesting behaviour in birds being published each year, much remains unknown.

@@

Several hypotheses have been proposed to explain the adaptive function of GNM (reviewed in @dubiec2013; @scott-baumann2015) First, the courtship hypothesis suggests that (mostly) males use GNM as visual or olfactory signals to attract females and/or advertise nest site occupancy to other males (e.g., @eens1993; @gwinner1997; @brouwer2004)), particularly in cavity-nesting species where visual cues inside the nest are limited. The courtship hypothesis traces back to the early natural history observations by @kessel1957, who noted that male European Starlings (Sturnus vulgaris) carried fresh plant material to their nests while singing and displaying to females. Subsequent studies in both European and Spotless Starlings (Sturnus unicolor) supported this view that GNM may serve as a sexually selected signal of male quality by showing that females often prefer nests with fresh GNM, and that unpaired males conspicuously deliver GNM during courtship, stopping right after the onset of egg laying [@brouwer2004; @veiga2006], and increasing this behaviour in the presence of potential mates [@eens1993] The courtship hypothesis was later extended to explain GNM addition solely by females in Blue Tits, therefore suggesting that GNM might serve as a sexually selected signal of female quality [@tomás2013].

Second, the most widely tested hypothesis is the nest protection hypothesis, which proposes that the volatile compounds in GNM reduce the amount of parasites and pathogens in the nests, enhancing offspring survival. This hypothesis traces back to a comparative study in raptors (Falconiformes) that provided indirect evidence for it by comparing nest reuse, diet composition and GNM presence [@wimberger1984]. Soon after, a preference for GNM containing insect-repellent and antimicrobial volatiles was experimentally confirmed in European Starlings, therefore confirming two of the three key criteria for evaluating the nest protection hypothesis (i.e., selective plant use, bioactive chemical properties and demonstrable adaptive benefits; Clark and Mason 1988). Observational and experimental studies on raptors [@ontiveros2007; @dykstra2009] and passerines [@lafuma2001; @dawson2004; @mennerat2009b; @quiroga2012] have shown mixed results. In some cases, GNM was shown to improve offspring condition without reducing the amount of parasites in the nests [@gwinner2000], which led to the development of the drug hypothesis. This alternative hypothesis suggests that the volatiles present in GNM directly enhance offspring immune function and/or health independent of parasite or pathogen control [@gwinner2000; @mennerat2009a]. Thus, while both the nest protection and drug hypotheses suggest an adaptive benefit of GNM by enhancing offspring survival, the drug hypothesis emphasizes a pharmacological mechanism, whereas the nest protection hypothesis emphasizes parasite and pathogen repellency. Importantly, these two hypotheses are not mutually exclusive unless the improvement in offspring health can be proved to be unrelated to a reduction in parasite or pathogen load, something difficult to achieve.

@@

We extracted the means, variances and sample sizes for each control-treatment pairwise comparison. Depending on the experimental design of each study, controls included non-aromatic materials such as grass or moss, lesser amounts of aromatic material than the treatments, no additional material added, or removal of material added by birds; whereas treatments consisted of aromatic materials added by authors, birds, or both. Those comparisons were coded either as non-aromatic GNM vs. aromatic GNM, or no added GNM vs. aromatic GNM, in our moderator "experimental design". However, there was a third type of comparison used in some studies comparing nests with experimentally added non-aromatic GNM to nests without any added GNM or in which GNM added by the birds was removed by the authors. In this type of comparison, the non-aromatic GNM was considered the treatment, and we coded these `{r} nrow(lnRR_comp3)` and `{r} nrow(SMDH_comp3)` effect sizes, respectively, as non-aromatic GNM vs. no added GNM in our moderator "experimental design". Importantly, our conclusions were robust to these effect sizes (\@Table SX), so we kept them in our dataset, and the moderator "experimental design" was used for subsequent meta-regressions. In studies where three groups were compared (n = `{r} nrow(multiple_comparision)`), we extracted all possible pairwise comparisons.

For each study, we also extracted information related to the study subjects (species of birds, geographical location of the study population), the methodology of the study (plant species used as treatment and control, fitness proxy measured) and the source from which we extracted data (e.g., figures, tables, raw data files, supplementary materials). We categorised the fitness proxies measured into 6 broad categories: physiology, morphology, reproduction, behaviour, parasite and pathogenic load, and phenology. For studies testing whether GNM affects the number of parasites or pathogens found in offspring and/or nests, we categorised them into arthropods or microorganisms. For a complete list of variables extracted, see @figsupp-variables . We did not include some traits reported in the original studies (k = `{r} nrow(excluded_proxies)` effect sizes, n = `{r} length(unique(excluded_proxies$paper_ID))` studies) in our analyses because it was unclear from the study and our own judgement how they related to fitness (see @figsupp-excluded-proxies).

To obtain missing data (e.g., non-reported results, missing sample sizes), we contacted four corresponding authors. All four authors responded to our email, but only two provided the data as requested; one did not have the data any more, and one did not give a conclusive response. Additionally, to locate grey literature (i.e., unpublished data), we wrote to the corresponding author(s) of all the (n = `{r} (nrow(selected_abstract_screening) - nrow(excluded_fulltext))` articles that passed full-text screening to ask for any potential unpublished data that could be included in our meta-analysis using a standardised email template (@figsupp-email). One author provided data for an unpublished study on Blue Tits (see description in @figsupp-unpublished-study) and one of us (MO), provided data for an unpublished study on Common Buzzards (Buteo buteo).

@@

Using the means, standard deviations and sample sizes for each control-treatment pairwise comparison, we calculated two effect sizes: a log response ratio (lnRR) and a standardised mean difference with heteroskedasticity correction (SMDH), using the escalc function (measure = ROM and SMDH respectively, vtype = "LS") in the R package 'metafor' v4.8-0 [@metafor]. We specified the treatment group as the numerator and the control group as the denominator for lnRR, and treatment minus control for SMDH, such that positive values indicate a 'fitness' increase in the treatment group. In cases such as measures of parasite load or prevalence, laying or hatching dates, courtship time, or scab score, the effect size sign was inverted prior to analysis to ensure that a positive value consistently meant an increase in fitness regardless of fitness proxy (i.e., to ensure comparability across effect sizes). SMDH could only be calculated when there were enough data points in both treatment and control groups to estimate the pooled variance; specifically, when the degrees of freedom (i.e., n1+n2−2) was greater than 1. There were `{r} as.character(english::english(nrow(low_sample), 1, 1))` effect sizes from `{r} as.character(english::english(length(unique(low_sample$paper_ID)), 1, 1))`) study that did not fulfil that requirement, which we then excluded entirely from the dataset.

For the few cases for which we could not extract means and variances but inferential statistics such as t and F values, we estimated SMD values (k = `{r} nrow(inferential_only)`, n = `{r} length(unique(inferential_only$paper_ID))`) using the equations provided by @thehand2019. For dichotomous variables (e.g., number of successful nests, k = `{r} nrow(contingency_table)`, n = `{r} length(unique(contingency_table$paper_ID))`), we calculated the log odds ratio (lnOR) using the 'escalc' function (measure=OR) from 'metafor' and subsequently converted it to SMD using equations from @thehand2019. Since lnRR cannot be calculated from inferential statistics, or dichotomous variables, or for comparisons including zero or negative values (k = `{r} nrow(negative_lnRR)`), the number of effect sizes included in the lnRR (k = `{r} nrow(dataset_lnRR)`, n = `{r} length(unique(dataset_lnRR$paper_ID))`) subset is smaller than that of SMDH (k = `{r} nrow(dataset_SMDH)`, n = `{r} length(unique(dataset_SMDH$paper_ID))`).

@@

We performed all analyses in R version 4.5.1 [@R-version]. All multilevel meta-analytical models and meta-regressions were fitted using 'metafor' v.4.8-0 [@metafor]. Sampling variances were modelled as variance-covariance matrices assuming a 0.5 correlation (ρ) between sampling variances from the same study [@noble2017] using the vcalc function from 'metafor' (See Supplementary Appendix @figsupp-correlation for sensitivity analyses using different values of *ρ*).

We originally considered the following levels of non-independence for our analyses: multiple effect sizes collected from  (a) same study (Paper ID), (b) bird species (Species ID) and (c) geographical locations (population ID), (d) sub-populations within the same location (Experiment ID; e.g., birds studied at the same location for multiple years or birds studied at the same location separated by distance), (e) groups of individuals within a population (Group ID; e.g., males and females in a population, first clutch and second clutch) were represented in our data several times, and (f) repeated measurement of the same trait within the same individuals (Repeated trait ID; e.g., chick mass on day 7, 14 and 21).

Unless stated otherwise, we fitted the following random effects in all our models to account for non-independence: Paper ID, Group ID, Species ID, Population ID and an observation-level random effect (Observation ID). Our random effect structure deviated from our pre-registration since, initially, we had pre-registered the inclusion of Experiment ID and Repeated trait ID but not Population ID. However, Experiment ID and Group ID were almost indistinguishable, so we decided to leave out the one with lesser resolution, i.e. Experiment ID. Additionally, since few effect sizes were measured repeatedly, making Repeated trait ID almost identical to the observation-level random effect Observation ID, we excluded it (@figsupp-random-effect). Last, we decided to add Population ID since we noticed after extracting the data but before any analyses that several studies were performed in the same geographical locations.

@@

To estimate the effect of GNM on fitness across birds, we used two effect sizes: lnRR and SMD(H). Thus, throughout, we fitted two separate sets of model for each effect size. First, we fitted multilevel intercept-only meta-analytical model to test the overall effect, its uncertainty and heterogeneity. We used the `rma.mv` function from the 'metafor' package using the 'REML' method and a *t* distribution for calculating the test statistics and confidence intervals for the moderators. We also conducted a sensitivity analysis including some traits for which the relationship with fitness is unclear or debated in the literature (e.g., sex ratio, testosterone levels), to test the robustness of our results (@figsupp-MLMA-excluded; k = `{r} nrow(SAonly_proxies)` , n = `{r} length(unique(SAonly_proxies$paper_ID))`). For all intercept-only models, we estimated the total amount of absolute heterogeneity ($σ^2$; the total variation not explained by sampling error), its sources or relative heterogeneity ($I^2$), mean-standardised metric (CVH2) and variance-mean-standardised metric of heterogeneity (M2) following the pluralistic approach suggested by @yang2024a .

To test the relative importance of the two main adaptive hypotheses suggested for GNM, that is, the courtship and the parental care hypothesis, which combines the nest protection and drug hypotheses, we ran a multilevel meta-regression including the moderator 'hypothesis' (levels: courtship, parental care, both). For this and all meta-regressions, we estimated $R_{marginal}^2$,which corresponds to the percentage of heterogeneity explained by the moderators [@nakagawa2012]. Unless stated otherwise, all meta-regressions correspond to unimoderator meta-regressions.

@@

In addition, we performed X additional multilevel meta-regressions intended to explore the potential threats to the internal validity of each study on our conclusions (i.e., Risk of Bias or RoB) following recommendations in Culina et al. (2025). ...

@@

Both lnRR and SMD(H) intercept-only multilevel meta-analyses showed an average increase in fitness outcomes in nests with an experimental increase in GNM compared to control nests; however, the overall effect was only statistically significant for SMD(H).

@@

summary(lnRR_design_pc)
car::linearHypothesis(lnRR_design_pc, rbind(c(0,1,-1)))
summary(SMDH_design_pc)
car::linearHypothesis(SMDH_design_pc, rbind(c(0,1,-1)))
```

@@

Third, although none of the species-specific mean estimates were statistically significant, most were positive [[\@figsupp-MR](C-D); @tabmain-MLMR], and bird species identity explained a non-negligible amount of heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_birds)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_birds)*100,1))[1]`%). Fourth, most fitness trait-specific mean estimates were positive, but only morphology was statistically significant and reproduction marginally so, both only for SMD(H) (SMDH~morphology~ = `{r} summary_SMDH_trait$estimate[2]`, 95% CI: `{r} summary_SMDH_trait$CI[2]`, 95% PI: `{r} summary_SMDH_trait$PI[2]`, p-value = `{r} summary_SMDH_trait$p_value[2]`, k = `{r} summary_SMDH_trait$k[2]`, n = `{r} summary_SMDH_trait$n[2]`; SMDH~reproduction~ = `{r} summary_SMDH_trait$estimate[6]`, 95% CI: `{r} summary_SMDH_trait$CI[6]`, 95% PI: `{r} summary_SMDH_trait$PI[6]`, p-value = `{r} summary_SMDH_trait$p_value[6]`, k = `{r} summary_SMDH_trait$k[6]`, n = `{r} summary_SMDH_trait$n[6]`; @figsupp-MR; @tabmain-MLMR) and trait type accounted for small amount of the heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_trait)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_trait)*100,1))[1]`%). Last, for the subset of outcomes relating to pathogen and parasitic load, estimates for arthropods and micro-organisms did not differed statistically signficantly (\@figsupp-MR; @tabmain-MLMR) and the type of parasite or pathogen explained very little heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_parasite)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_parasite)*100,1))[1]`%).

@@

>>>>

Nest and nest-building are fundamental across a diverse range of animals, including arthropods (e.g., termites, spiders), mammals (e.g., platypus, rodents), and perhaps more famously, birds. Nests provide structural support for eggs, parents, and offspring, and protect them against predators and environmental pressures, and are thus key to reproductive success and survival [@hansell2000; @mainwaring2014; @deeming2023]. While most birds use organic plant materials such as twigs, branches, grass, and moss to build their nests, some incorporate inorganic items, including mud, stones, feathers, snake skins and even anthropogenic materials like plastic and cigarette butts [@sergio2011; @suárez-rodríguez2013; @jagiello2023; @sheard2024]. The latter are considered non-structural additions and suggest that nests are more than just physical shelters; they can be complex tools shaped by both natural and sexual selection for multiple functions. Indeed, nests can play a key role in sexual display, and features such as size have been shown to act as sexually selected signals of parental quality in several bird species [e.g., @moreno1994; @hoi1996], even interspecifically by brood parasites [@soler1995]. Nest-building behaviour is increasingly recognized as an expression of the extended phenotype (reviewed by @moreno2012), and phylogeny has been shown to be an important factor explaining both mean and within-species variation in nest morphology [@perez2023]. Yet, despite hundreds of scientific articles studying nests and nesting behaviour in birds being published each year, much remains unknown.

@@

Several hypotheses have been proposed to explain the adaptive function of GNM (reviewed in @dubiec2013; @scott-baumann2015). First, the courtship hypothesis suggests that (mostly) males use GNM as visual or olfactory signals to attract females and/or advertise nest site occupancy to other males (e.g., @eens1993; @gwinner1997; @brouwer2004), particularly in cavity-nesting species where visual cues inside the nest are limited. The courtship hypothesis traces back to the early natural history observations by @kessel1957, who noted that male European Starlings (Sturnus vulgaris) carried fresh plant material to their nests while singing and displaying to females. Subsequent studies in both European and Spotless Starlings (Sturnus unicolor) supported this view that GNM may serve as a sexually selected signal of male quality by showing that females often prefer nests with fresh GNM, and that unpaired males conspicuously deliver GNM during courtship, stopping right after the onset of egg laying [@brouwer2004; @veiga2006], and increasing this behaviour in the presence of potential mates [@eens1993]. The courtship hypothesis was later extended to explain GNM addition solely by females in Blue Tits, therefore suggesting that GNM might serve as a sexually selected signal of female quality [@tomás2013].

Second, the most widely tested hypothesis is the nest protection hypothesis, which proposes that the volatile compounds in GNM reduce the amount of parasites and pathogens in the nests, enhancing offspring survival. This hypothesis traces back to a comparative study in raptors (Falconiformes) that provided indirect evidence for it by comparing nest reuse, diet composition and GNM presence [@wimberger1984]. Soon after, a preference for GNM containing insect-repellent and antimicrobial volatiles was experimentally confirmed in European Starlings and suggested three key criteria for evaluating the nest protection hypothesis [i.e., selective plant use, bioactive chemical properties and demonstrable adaptive benefits; @clark1985]. Observational and experimental studies on raptors [@ontiveros2007; @dykstra2009] and passerines [@lafuma2001; @dawson2004; @mennerat2009b; @quiroga2012] have shown mixed results. In some cases, GNM was shown to improve offspring condition without reducing the amount of parasites in the nests [@gwinner2000], which led to the development of the drug hypothesis. This alternative hypothesis suggests that the volatiles present in GNM directly enhance offspring immune function and/or health independent of parasite or pathogen control [@gwinner2000; @mennerat2009a]. Thus, while both the nest protection and drug hypotheses suggest an adaptive benefit of GNM by enhancing offspring survival, the drug hypothesis emphasizes a pharmacological mechanism, whereas the nest protection hypothesis emphasizes parasite and pathogen repellency. Importantly, these two hypotheses are not mutually exclusive unless the improvement in offspring health can be proved to be unrelated to a reduction in parasite or pathogen load, something difficult to achieve.

@@

We extracted the means, variances and sample sizes for each control-treatment pairwise comparison. Depending on the experimental design of each study, controls included non-aromatic materials such as grass or moss, lesser amounts of aromatic material than the treatments, no additional material added, or removal of material added by birds; whereas treatments consisted of aromatic materials added by authors, birds, or both. Those comparisons were coded either as non-aromatic GNM vs. aromatic GNM, or no added GNM vs. aromatic GNM, in our moderator "experimental design". However, there was a third type of comparison used in some studies comparing nests with experimentally added non-aromatic GNM to nests without any added GNM or in which GNM added by the birds was removed by the authors. In this type of comparison, the non-aromatic GNM was considered the treatment, and we coded these `{r} nrow(lnRR_comp3)` and `{r} nrow(SMDH_comp3)` effect sizes, respectively, as non-aromatic GNM vs. no added GNM in our moderator "experimental design". Importantly, our conclusions were robust to the exclusion of these effect sizes (\@Table SX), so we kept them in our dataset, and the moderator "experimental design" was used for subsequent meta-regressions. In studies where three groups were compared (n = `{r} nrow(multiple_comparision)`), we extracted all possible pairwise comparisons, and accounted for (shared-group) non-independence in our analyses (more below).

For each study, we also extracted information related to the study subjects (species of birds, geographical location of the study population), the methodology of the study (plant species used as treatment and control, fitness proxy measured) and the source from which we extracted data (e.g., figures, tables, raw data files, supplementary materials). We categorised the fitness proxies measured into 6 broad categories: physiology, morphology, reproduction, behaviour, parasite and pathogenic load, and phenology. For studies testing whether GNM affects the number of parasites or pathogens found in offspring and/or nests, we categorised them into arthropods or microorganisms. For a complete list of variables extracted, see @figsupp-variables . During data extraction, we encountered some traits (e.g., white-blood-cell ratio, parent mass; `{r} nrow(excluded_proxies)` effect sizes, n = `{r} length(unique(excluded_proxies$paper_ID))` studies) that we did not include in any of our analyses because it was unclear from the study and our own judgement how these traits related to fitness (see @figsupp-excluded-proxies for the list and reasons of exclusion).

To obtain missing data (e.g., non-reported results, missing sample sizes), we contacted four corresponding authors. All four authors responded to our email, but only two provided the data as requested; one did not have the data any more, and one did not give a conclusive response. Additionally, to locate grey literature (i.e., unpublished data), we wrote to the corresponding author(s) of all the (n = `{r} (nrow(selected_abstract_screening) - nrow(excluded_fulltext))` articles that passed full-text screening to ask for any potential unpublished data that could be included in our meta-analysis using a standardised email template (@figsupp-email). One author provided data for an unpublished study on Blue Tits (see description in @figsupp-unpublished-study) and one of us (MO) provided data for an unpublished study on Common Buzzards (*Buteo buteo*).

@@

Using the means, standard deviations and sample sizes for each control-treatment pairwise comparison, we calculated two effect sizes: a log response ratio (lnRR) and a standardised mean difference with heteroskedasticity correction (SMDH), using the 'escalc' function (measure = ROM and SMDH, respectively, vtype = "LS") in the R package 'metafor' v4.8-0 [@metafor]. We specified the treatment group as the numerator and the control group as the denominator for lnRR, and treatment minus control for SMDH, such that positive values indicate a 'fitness' increase in the treatment group. In cases such as measures of parasite load or prevalence, laying or hatching dates, courtship time, or scab score, the effect size sign was inverted before analysis to ensure that a positive value consistently meant an increase in fitness regardless of fitness proxy (i.e., to ensure comparability across effect sizes). SMDH could only be calculated when there were enough data points in both treatment and control groups to estimate the pooled variance; specifically, when the degrees of freedom (i.e., n1+n2−2) was greater than 1. There were `{r} as.character(english::english(nrow(low_sample), 1, 1))` effect sizes from `{r} as.character(english::english(length(unique(low_sample$paper_ID)), 1, 1))`) study that did not fulfil that requirement, which we then excluded entirely from the dataset.

For the few cases for which we could not extract means and variances but inferential statistics such as t and F values, we estimated SMD values (k = `{r} nrow(inferential_only)`, n = `{r} length(unique(inferential_only$paper_ID))`) using the equations provided by @thehand2019. For dichotomous variables (e.g., number of successful nests, k = `{r} nrow(contingency_table)`, n = `{r} length(unique(contingency_table$paper_ID))`), we calculated the log odds ratio (lnOR) using the 'escalc' function (measure=OR) from 'metafor' and subsequently converted it to SMD using equations from @thehand2019. Since lnRR cannot be calculated from inferential statistics, dichotomous variables, or traits including zero or negative values (k = `{r} nrow(negative_lnRR)`), the number of effect sizes included in the lnRR (k = `{r} nrow(dataset_lnRR)`, n = `{r} length(unique(dataset_lnRR$paper_ID))`) subset is smaller than that of SMDH (k = `{r} nrow(dataset_SMDH)`, n = `{r} length(unique(dataset_SMDH$paper_ID))`).

@@

We performed all analyses in R v4.5.1 [@R-version]. All multilevel meta-analytical models and meta-regressions were fitted using 'metafor' v.4.8-0 [@metafor]. Sampling variances were modelled as variance-covariance matrices assuming a 0.5 correlation (ρ) between sampling variances from the same study [@noble2017] using the vcalc function from 'metafor' (See Supplementary Appendix @figsupp-correlation for sensitivity analyses using different values of *ρ*).

We originally considered the following levels of non-independence for our analyses: multiple effect sizes collected from  (a) same study (Paper ID), (b) bird species (Species ID) and (c) geographical locations (population ID), (d) sub-populations within the same location (Experiment ID; e.g., birds studied at the same location for multiple years), (e) groups of individuals within a population (Group ID; e.g., males and females in a population, first clutch and second clutch) were represented in our data several times, and (f) repeated measurement of the same trait within the same individuals (Repeated trait ID; e.g., chick mass on day 7, 14 and 21).

Unless stated otherwise, we fitted the following random effects in all our models to account for non-independence: Paper ID, Group ID, Species ID, Population ID and an observation-level random effect (Observation ID). Our random effect structure deviated from our pre-registration since, initially, we had pre-registered the inclusion of Experiment ID and Repeated trait ID but not Population ID. However, Experiment ID and Group ID were almost indistinguishable, so we decided to leave out the one with lesser resolution, i.e. Experiment ID. Additionally, few effect sizes were measured repeatedly, making Repeated trait ID almost identical to Observation ID (@figsupp-random-effect), so we excluded Repeated trait ID. Last, we decided to add Population ID since we noticed after extracting the data, but before any analyses, that several studies were performed in the same geographical locations.

@@

To estimate the effect of GNM on fitness across birds, we used two effect sizes: lnRR and SMD(H). Thus, throughout, we fitted two separate sets of model for each effect size. First, we fitted multilevel intercept-only meta-analytical model to test the overall effect, its uncertainty and heterogeneity. We used the `rma.mv` function from the 'metafor' package using the 'REML' method and a *t* distribution for calculating the test statistics and confidence intervals for the moderators. We also conducted a sensitivity analysis including some traits for which the relationship with fitness is debated in the literature (e.g., sex ratio, testosterone levels), to test the robustness of our results (@figsupp-MLMA-excluded; k = `{r} nrow(SAonly_proxies)` , n = `{r} length(unique(SAonly_proxies$paper_ID))`). For all intercept-only models, we estimated the total amount of absolute heterogeneity ($σ^2$; the total variation not explained by sampling error), its sources or relative heterogeneity ($I^2$), mean-standardised metric (CVH2) and variance-mean-standardised metric of heterogeneity (M2) following the pluralistic approach suggested by @yang2024a .

To test the relative importance of the two main adaptive hypotheses suggested for GNM, that is, the courtship and the parental care hypothesis, which combines the nest protection and drug hypotheses, we ran a multilevel meta-regression including the moderator 'hypothesis' (levels: courtship, parental care, both). For this and all meta-regressions, we estimated $R_{marginal}^2$, which corresponds to the percentage of heterogeneity explained by the moderators [@nakagawa2012]. Unless stated otherwise, all meta-regressions correspond to unimoderator meta-regressions.

@@

In addition, we performed three additional pre-registered multilevel meta-regressions intended to explore the potential threats to the internal validity of each study on our conclusions (i.e., Risk of Bias or RoB) following recommendations in @culina2025 . We evaluated the effects of (1) blinding (whether outcome assessors were blinded to the experimental group), (2) random assignment to treatment, and (3) partial reporting (whether all measured outcomes were completely reported or only a subset). For each RoB category, we coded studies (levels: yes, no) and included each category as a moderator in a meta-regression separately for both lnRR and SMD(H).

@@

Both lnRR and SMD(H) intercept-only multilevel meta-analyses showed an average increase in fitness estimates in nests with an experimental increase in GNM compared to control nests; however, the overall effect was only statistically significant for SMD(H).

@@

summary(lnRR_design_pc)
car::linearHypothesis(lnRR_design_pc, rbind(c(0,1,-1)))
summary(SMDH_design_pc)
car::linearHypothesis(SMDH_design_pc, rbind(c(0,1,-1)))

# Now loading the models for RoB Meta-regressions

# Loading models for blinding performed or not
lnRR_blinding<-readRDS(here::here("model/lnRR_blinding.rds"))
SMDH_blinding<-readRDS(here::here("model/SMDH_blinding.rds"))

summary_lnRR_blinding<-extractor_metareg(lnRR_blinding,"blinding")
summary_SMDH_blinding<-extractor_metareg(SMDH_blinding,"blinding")

# Loading models for random sampling assignment performed or not
lnRR_random_assignment<-readRDS(here::here("model/lnRR_random_assignment.rds"))
SMDH_random_assignment<-readRDS(here::here("model/SMDH_random_assignment.rds"))

summary_lnRR_random_assignment<-extractor_metareg(lnRR_random_assignment,"random_assignment")
summary_SMDH_random_assignment<-extractor_metareg(SMDH_random_assignment,"random_assignment")

# Loading models for missing data present or not
lnRR_missing_data<-readRDS(here::here("model/lnRR_missing_data.rds"))
SMDH_missing_data<-readRDS(here::here("model/SMDH_missing_data.rds"))

summary_lnRR_missing_data<-extractor_metareg(lnRR_missing_data,"missing_data")
summary_SMDH_missing_data<-extractor_metareg(SMDH_missing_data,"missing_data")

@@

Third, although none of the species-specific mean estimates were statistically significant, most were positive [[\@figsupp-MR](C-D); @tabmain-MLMR], and bird species identity explained a non-negligible amount of heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_birds)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_birds)*100,1))[1]`%). Fourth, most fitness trait-specific mean estimates were positive, but only morphology was statistically significant only for SMD(H) (SMDH~morphology~ = `{r} summary_SMDH_trait$estimate[2]`, 95% CI: `{r} summary_SMDH_trait$CI[2]`, 95% PI: `{r} summary_SMDH_trait$PI[2]`, p-value = `{r} summary_SMDH_trait$p_value[2]`, k = `{r} summary_SMDH_trait$k[2]`, n = `{r} summary_SMDH_trait$n[2]`; @figsupp-MR; @tabmain-MLMR) and trait type accounted for small amount of the heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_trait)*100,1))[1]`% and $R^2_{marginal(SMD(H))}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_trait)*100,1))[1]`%). Fifth, for the subset of outcomes relating to pathogen and parasitic load, estimates for arthropods and micro-organisms did not differed statistically significantly (\@figsupp-MR; @tabmain-MLMR) and the type of parasite or pathogen explained very little heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_parasite)*100,1))[1]`% and $R^2_{marginal(SMD(H))}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_parasite)*100,1))[1]`%).

Lastly, the risk of bias categories we evaluated (i.e., blinding, random assignment to each treatment and partial reporting) were not statistically significant for lnRR ([@tabsupp-SA-RoB]). SMD(H) was positive and statistically significant for studies without blinding (SMD(H)~No Blinding~ = `{r} summary_SMDH_blinding$estimate[1]`, 95% CI: `{r} summary_SMDH_blinding$CI[1]`, 95% PI: `{r} summary_SMDH_blinding$PI[1]`, p-value = `{r} summary_SMDH_blinding$p_value[1]`, k = `{r} summary_SMDH_blinding$k[1]`, n = `{r} summary_SMDH_blinding$n[1]`; @tabsupp-SA-RoB), with random assignment to treatment (SMD(H)~Random Assignment~ = `{r} summary_SMDH_random_assignment$estimate[2]`, 95% CI: `{r} summary_SMDH_random_assignment$CI[2]`, 95% PI: `{r} summary_SMDH_random_assignment$PI[2]`, p-value = `{r} summary_SMDH_random_assignment$p_value[2]`, k = `{r} summary_SMDH_random_assignment$k[2]`, n = `{r} summary_SMDH_random_assignment$n[2]`; @tabsupp-SA-RoB) and where no data was missing (SMD(H)~No Partial Reporting~ = `{r} summary_SMDH_missing_data$estimate[1]`, 95% CI: `{r} summary_SMDH_missing_data$CI[1]`, 95% PI: `{r} summary_SMDH_missing_data$PI[1]`, p-value = `{r} summary_SMDH_missing_data$p_value[1]`, k = `{r} summary_SMDH_missing_data$k[1]`, n = `{r} summary_SMDH_missing_data$n[1]`; @tabsupp-SA-RoB). Notably, other levels of each category included fewer than four studies, indicating that the observed significance likely reflects the higher statistical power in the larger sample level rather than any systematic bias. All the three categories for risk of bias explained very little heterogeneity ($R^2_{marginal(lnRR)}$\> `{r} as.double(round(orchaRd::r2_ml(lnRR_blinding)*100,1))[1]`%; $R^2_{marginal(SMD(H))}$\> `{r} as.double(round(orchaRd::r2_ml(SMDH_random_assignment)*100,1))[1]`%).

@@

```{r}
lnRR_small_study<-readRDS(here::here("model/lnRR_small_study.rds"))
SMDH_small_study<-readRDS(here::here("model/SMDH_small_study.rds"))

lnRR_decline_eff<-readRDS(here::here("model/lnRR_decline_eff.rds"))
SMDH_decline_eff<-readRDS(here::here("model/SMDH_decline_eff.rds"))
```

We found no clear evidence for publication bias in the published literature [@figsupp-pub-bias]. There was no evidence for small-study effect for lnRR i.e. studies with smaller sample sizes did not have larger effect size estimates (lnRR~intercept~= `{r} sprintf("%.3f", coef(lnRR_small_study)[1])`, 95% CI: \[`{r} sprintf("%.3f", lnRR_small_study$ci.lb[1])`, `{r} sprintf("%.3f", lnRR_small_study$ci.ub[1])`\], p-value = `{r} sprintf("%.3f", lnRR_small_study$pval[1])`, lnRR~slope~ = `{r} sprintf("%.3f", coef(lnRR_small_study)[2])`, 95% CI: \[`{r} sprintf("%.3f", lnRR_small_study$ci.lb[2])`; `{r} sprintf("%.3f", lnRR_small_study$ci.ub[2])`\], p-value = `{r} sprintf("%.3f", lnRR_small_study$pval[2])`, k = `{r} lnRR_small_study$k`, n = `{r} length(lnRR_small_study$s.levels.f[[1]])`, $R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_small_study)*100,1))[1]`%). There was a tendency for effect sizes to be larger when sample size was smaller but this relationship was not statistically significant (SMDH~slope~ = `{r} sprintf("%.3f", coef(SMDH_small_study)[2])`, 95% CI: \[`{r} sprintf("%.3f", SMDH_small_study$ci.lb[2])`, `{r} sprintf("%.3f", SMDH_small_study$ci.ub[2])`\], p-value = `{r} sprintf("%.3f", SMDH_small_study$pval[2])`, k = `{r} SMDH_small_study$k`, n = `{r} length(SMDH_small_study$s.levels.f[[1]])`, $R^2_{marginal(SMD(H))}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_small_study)*100,1))[1]`%). After adjusting for the funnelplot asymmetry (i.e., small study effect), the overall effect of GNM on fitness outcomes disappeared (SMDH~intercept~ = `{r} sprintf("%.3f", coef(SMDH_small_study)[1])`, 95% CI: \[`{r} sprintf("%.3f", SMDH_small_study$ci.lb[1])`, `{r} sprintf("%.3f", SMDH_small_study$ci.ub[1])`\], p-value = `{r} sprintf("%.3f", SMDH_small_study$pval[1])`).
